---
title: "Introduction to alluvialmatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to alluvialmatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
# devtools::load_all()
library(alluvialmatch)
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(ggforce)
library(igraph)
library(tibble)
```

We create a toy data frame that maps tissues (brain, stomach, heart, T cell, B cell) to clustering (1-4)
```{r}
df <- data.frame(
  tissue = c(
    "brain", "brain", "brain",
    "stomach", "stomach", "stomach", "stomach", "stomach", "stomach",
    "heart", "heart", "heart", "heart", "heart", "heart", "heart",
    "T cell", "T cell",
    "B cell", "B cell", "B cell", "B cell", "B cell", "B cell", "B cell", "B cell", "B cell"
  ),
  cluster = c(
    1, 1, 2,
    1, 2, 2, 2, 2, 2,
    1, 3, 3, 3, 3, 3, 3,
    4, 4,
    4, 4, 4, 4, 4, 4, 4, 4, 4
  )
)

head(df)
```

Group by tissue and cluster, counting the values in a new column value
```{r}
clus_df_gather <- df |>
    dplyr::mutate_if(is.numeric, function(x) factor(x, levels = as.character(sort(unique(x))))) |>
    dplyr::group_by_all() |>
    dplyr::count(name = "value")
gather_set_data(clus_df_gather, 1:2)

clus_df_gather
```


Plot with ggalluvial before sorting - notice especially how B cells cross down to cluster 4, and stomach crosses up to cluster 2 - we would like to straighten this out, and add colors to show corresponding groups
```{r}
p1 <- ggplot(data = clus_df_gather, aes(axis1 = tissue, axis2 = cluster, y = value)) +
    geom_alluvium() +
    geom_stratum() +
    scale_x_discrete(limits = c("tissue", "cluster"), expand = c(.2, .05)) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    theme_minimal()
p1
```
Implementing a greedy heuristic for weighted one-layer free problem
```{r}
clus_df_gather_sorted <- greedy_wolf(clus_df_gather, column1="tissue", column2="cluster", column_weights = "value")
clus_df_gather_sorted
```

Plot the alluvial plot with sorting and color matching
```{r}
p2 <- plot_alluvial(clus_df_gather, column1="tissue", column2="cluster", column_weights = "value")
# alternatively: p2 <- plot_alluvial(df, column1="tissue", column2="cluster")
p2
```

```{r}
sessioninfo::session_info()
```
