---
title: "Calling wompwomp from Inside ggalluvial"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Calling wompwomp from Inside ggalluvial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(wompwomp)
set.seed(43)
```

The `vignette("plotting")` workflow sorts and colours the data frame first and
hands the result to ggplot2. Development versions of **ggalluvial** can instead
call wompwomp from within the alluvial layers themselves, through the
`sort_strata` and `color_strata` parameters of `stat_stratum()`,
`stat_alluvium()` and `stat_flow()`. The parameter names an *engine*;
`"wompwomp"` runs `sort_to_uncross(column_method = "none")` for `sort_strata`
and `get_lode_clusters()` for `color_strata`, the latter exposed as the computed
variable `cluster`. `sort_strata` must be given the same value in every layer of
a plot, since each layer computes its own positions; wompwomp is still run only
once per panel and the layers share the result.

We use the same toy data as the other vignettes.

```{r}
df <- data.frame(
  tissue = c(
    "BRAIN", "BRAIN", "BRAIN",
    "STOMACH", "STOMACH", "STOMACH", "STOMACH", "STOMACH", "STOMACH",
    "HEART", "HEART", "HEART", "HEART", "HEART", "HEART", "HEART",
    "T CELL", "T CELL",
    "B CELL", "B CELL", "B CELL", "B CELL", "B CELL", "B CELL", "B CELL", "B CELL", "B CELL"
  ),
  cluster = c(
    1, 1, 2,
    1, 2, 2, 2, 2, 2,
    1, 3, 3, 3, 3, 3, 3,
    4, 4,
    4, 4, 4, 4, 4, 4, 4, 4, 4
  )
)

df |> 
  dplyr::group_by(tissue, cluster) |> 
  dplyr::summarize(value = dplyr::n()) |> 
  dplyr::ungroup() |> 
  dplyr::mutate(dplyr::across(c(tissue, cluster), as.character)) |> 
  print() -> clus_df_gather
```

## Sorting and colouring inside the layers

```{r, eval = utils::packageVersion("ggalluvial") > "0.12.6"}
clus_df_gather |>
  ggplot(aes(y = value, axis1 = tissue, axis2 = cluster)) +
  geom_alluvium(aes(fill = tissue), width = 1/12,
                sort_strata = "wompwomp") +
  geom_stratum(aes(fill = after_stat(cluster)), width = 1/12, color = "grey",
               sort_strata = "wompwomp", color_strata = "wompwomp") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)),
             sort_strata = "wompwomp") +
  scale_x_discrete(
    limits = c("tissue", "cluster"),
    expand = c(.05, .05)
  )
```

## Choosing and tuning an algorithm

To choose an algorithm other than the default, or to tune it, pass a list: the
`method` selects among `sort_to_uncross()`'s (or `get_lode_clusters()`'s)
methods, and the remaining named elements are that function's own arguments.
Everything else wompwomp exposes lives in `sort_to_uncross_options()` and
`get_lode_clusters_options()`, so those knobs travel in `options` --- passing
one directly is an error that names the constructor to use.

```{r, eval = FALSE}
# arguments of sort_to_uncross() / get_lode_clusters()
geom_stratum(sort_strata = list("wompwomp", method = "tsp", alpha = 3))
geom_stratum(color_strata = list("wompwomp", method = "advanced",
                                 resolution = 2))

# anything from the *_options() constructors goes through `options`
geom_stratum(sort_strata = list(
  "wompwomp",
  options = sort_to_uncross_options(weighted_metric = FALSE)
))
geom_stratum(color_strata = list(
  "wompwomp", method = "left",
  options = get_lode_clusters_options(cutoff = 0.3)
))
```

Note that `method = "tsp"` starts the solver from a random tour, so seed it
when a plot has to be redrawn identically; the default `"neighbornet"` is
deterministic.

## Axis order

The order of the *axes* stays outside the layers: position scales are trained
before statistical transformations run, so a `stat_*()` that permuted `x` would
move the strata out from under their axis labels. `ggalluvial::sort_axes()`
returns the recommended axis order instead, for use in
`scale_x_discrete(limits = )` --- exactly where the axis order is written
above.

```{r, eval = FALSE}
axis_order <- ggalluvial::sort_axes(
  clus_df_gather, axes = c("tissue", "cluster"), weight = "value",
  engine = "wompwomp"          # or list("wompwomp", method = "neighbornet")
)
```

It runs `sort_to_uncross()` with `column_method` set and reads off only the
column order. The strata are sorted while it searches, because candidate axis
orders are compared by the crossings they induce; that inner ordering defaults
to `"neighbornet"` and can be set with `strata_method`. Applying the result
re-sorts the strata against the new axis order, since which strata minimise
crossings depends on which axes end up adjacent.

Sorting the axes and then the strata is two passes over a problem
`sort_to_uncross(method = , column_method = )` optimises jointly, and the
column-ordering heuristic does not guarantee an improvement --- on some data it
returns an order no better than the one you started with. Check with
`compute_crossing_objective()` before adopting it, or call `sort_to_uncross()`
directly and plot the result the way `vignette("plotting")` does.

```{r}
sessionInfo()
```
