---
title: "Global optimum"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
        self_contained: false
vignette: >
  %\VignetteIndexEntry{4. Global optimum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
# devtools::load_all() # !!! erase

library(wompwomp)  #!!! uncomment
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(ggforce)
library(igraph)
library(tibble)
library(tidyr)

set.seed(42)
```

We create a toy data frame that maps tissues (BRAIN, STOMACH, HEART, T cell, B CELL) to clustering (1-5)
```{r}
df <- data.frame(
    tissue = c(
        1, 1, 1,
        2, 2, 2, 2, 2, 2,
        3, 3, 3, 3, 3, 3, 3,
        4, 4,
        5, 5, 5, 5, 5, 5, 5, 5, 5
    ),
    cluster = c(
        2, 2, 1,
        2, 1, 1, 1, 1, 1,
        2, 3, 3, 3, 3, 3, 3,
        4, 4,
        4, 4, 4, 4, 4, 4, 4, 4, 5
    )
)
column1 <- "tissue"
column2 <- "cluster"

head(df)
```

Implementing a greedy heuristic for weighted one-layer free problem
```{r}
clus_df_gather_sorted <- data_sort(df, column1 = "tissue", column2 = "cluster", sorting_algorithm = "greedy_wolf", fixed_column = "tissue")
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
wompwomp_objective_greedy_wolf <- determine_weighted_layer_free_objective(crossing_edges)
print(wompwomp_objective_greedy_wolf)
```

Plot the alluvial plot with sorting and color matching
```{r}
p1 <- plot_alluvial(clus_df_gather_sorted, column1 = "tissue", column2 = "cluster", column_weights = "value", sorting_algorithm = "greedy_wolf", fixed_column = "tissue", color_bands = TRUE)
p1
```

Repeat for all permutations of the right side, calculating objective for each one
```{r}
# Sample: assume clus_df_gather_sorted has factor 'cluster' with n levels
n <- nlevels(clus_df_gather_sorted$cluster)

# Get all n! permutations of labels 1 to n
perms <- gtools::permutations(n, n)

# Original cluster levels
orig_levels <- levels(clus_df_gather_sorted$col2_int)

# Loop over permutations
objective_minimum <- Inf
objectives <- numeric(nrow(perms))
for (i in 1:nrow(perms)) {
    new_labels <- perms[i, ] # as.character(perms[i, ])
    right_map <- setNames(new_labels, orig_levels)

    # Create a relabeled copy
    clus_permuted <- clus_df_gather_sorted %>%
        mutate(
            col2_int = right_map[as.character(col2_int)]
        )

    map_dict <- setNames(new_labels, orig_levels)
    clus_permuted_objective <- determine_crossing_edges(clus_permuted, column1 = "col1_int", column2 = "col2_int", return_weighted_layer_free_objective = TRUE)

    objectives[i] <- clus_permuted_objective

    if (clus_permuted_objective < objective_minimum) {
        clus_permuted_best <- clus_df_gather_sorted %>%
            mutate(cluster = factor(cluster, levels = orig_levels, labels = new_labels))
    }
}
```

```{r}
clus_permuted_best
```


```{r}
p_best_wolf <- plot_alluvial(clus_permuted_best, column1 = "col1_int", column2 = "col2_int", column_weights = "value", sorting_algorithm = "none", color_bands = TRUE)
p_best_wolf
```


```{r}
df_plot <- data.frame(
    perm_id = seq_along(objectives),
    objective = sort(objectives)
)

ggplot(df_plot, aes(x = perm_id, y = objective)) +
    geom_point() +
    geom_hline(yintercept = wompwomp_objective_greedy_wolf, linetype = "dashed", color = "blue") +
    annotate("text",
        x = Inf, y = wompwomp_objective_greedy_wolf, label = "wompwomp objective",
        hjust = 1.1, vjust = -0.5, color = "blue", size = 3
    ) +
    labs(x = "Permutation index", y = "Objective value") +
    theme_minimal()
```

Now do the same for WBLF
```{r}
clus_df_gather_sorted <- data_sort(df, column1 = "tissue", column2 = "cluster", sorting_algorithm = "greedy_wblf", random_initializations = 10)
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
wompwomp_objective_greedy_wblf <- determine_weighted_layer_free_objective(crossing_edges)
print(wompwomp_objective_greedy_wblf)
```

```{r}
p2 <- plot_alluvial(clus_df_gather_sorted, column1 = "tissue", column2 = "cluster", column_weights = "value", sorting_algorithm = "greedy_wblf", color_bands = TRUE)
p2
```

Note that, above, we would reach the optimal if we moved up the cyan's 1 notch up

```{r}
# Get levels
left_levels <- levels(clus_df_gather_sorted$col1_int)
right_levels <- levels(clus_df_gather_sorted$col2_int)

n_left <- length(left_levels)
n_right <- length(right_levels)

# Get all permutations of each side
left_perms <- gtools::permutations(n_left, n_left)
right_perms <- gtools::permutations(n_right, n_right)

# Initialize tracking variables
objective_minimum <- Inf
objectives <- matrix(NA, nrow = nrow(left_perms), ncol = nrow(right_perms))
best_left_map <- NULL
best_right_map <- NULL

# Loop through all left and right permutations
for (i in 1:nrow(left_perms)) {
    # print(i)
    left_new_labels <- left_perms[i, ]
    left_map <- setNames(left_new_labels, left_levels)

    for (j in 1:nrow(right_perms)) {
        right_new_labels <- right_perms[j, ]
        right_map <- setNames(right_new_labels, right_levels)

        # Apply mappings (e.g., to integer relabels)
        clus_permuted <- clus_df_gather_sorted %>%
            mutate(
                col1_int = left_map[as.character(col1_int)],
                col2_int = right_map[as.character(col2_int)]
            )

        # Compute objective
        obj <- determine_crossing_edges(
            clus_permuted,
            column1 = "col1_int",
            column2 = "col2_int",
            return_weighted_layer_free_objective = TRUE
        )

        objectives[i, j] <- obj

        # Update minimum if needed
        if (obj < objective_minimum) {
            objective_minimum <- obj
            clus_permuted_best <- clus_permuted
            best_left_map <- left_map
            best_right_map <- right_map
        }
    }
}
```

```{r}
p_best_wblf <- plot_alluvial(clus_permuted_best, column1 = "col1_int", column2 = "col2_int", column_weights = "value", sorting_algorithm = "none", color_bands = TRUE)
p_best_wblf
```

Let's compare tsp too
```{r}
clus_df_gather_sorted_tsp <- 0  # a default
```

```{r}
clus_df_gather_sorted_tsp <- data_sort(df, column1 = "tissue", column2 = "cluster", sorting_algorithm = "tsp")
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted_tsp, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
tsp_objective_minimum <- determine_weighted_layer_free_objective(crossing_edges)
print(tsp_objective_minimum)

plot_alluvial(clus_df_gather_sorted_tsp, column1 = "tissue", column2 = "cluster", column_weights = "value", sorting_algorithm = "none", color_bands = TRUE)
```

```{r}
df_plot <- data.frame(
    perm_id = seq_along(objectives),
    objective = sort(objectives)
)

ggplot(df_plot, aes(x = perm_id, y = objective)) +
    geom_point() +
    geom_hline(yintercept = wompwomp_objective_greedy_wblf, linetype = "dashed", color = "blue") +
    annotate("text",
        x = Inf, y = wompwomp_objective_greedy_wblf, label = "greedy WBLF objective",
        hjust = 1.1, vjust = -0.5, color = "blue", size = 3
    ) +
    geom_hline(yintercept = tsp_objective_minimum, linetype = "dashed", color = "orange") +
    annotate("text",
        x = Inf, y = tsp_objective_minimum, label = "tsp objective",
        hjust = 1.1, vjust = -0.5, color = "orange", size = 3
    ) +
    labs(x = "Permutation index", y = "Objective value") +
    theme_minimal()
```

```{r}
sessioninfo::session_info()
```
