---
title: "Introduction to alluvialmatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to alluvialmatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
# devtools::load_all()
library(alluvialmatch)
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(ggforce)
library(igraph)
library(tibble)
```

```{r}
library(reticulate)
conda_env <- "seurat_vs_scanpy"
Sys.setenv(RETICULATE_PYTHON = paste("/Users/joeyrich/miniconda3/envs", conda_env, "bin/python3.9", sep = "/"))
use_condaenv("/Users/joeyrich/miniconda3/envs/seurat_vs_scanpy/bin/python3.9", required = TRUE)
```

We create a toy data frame that maps tissues (BRAIN, STOMACH, HEART, T cell, B CELL) to clustering (1-5)
```{r}
df <- data.frame(
  tissue = c(
    1, 1, 1,
    2, 2, 2, 2, 2, 2,
    3, 3, 3, 3, 3, 3, 3,
    4, 4,
    5, 5, 5, 5, 5, 5, 5, 5, 5
  ),
  cluster = c(
    2, 2, 1,
    2, 1, 1, 1, 1, 1,
    2, 3, 3, 3, 3, 3, 3,
    4, 4,
    4, 4, 4, 4, 4, 4, 4, 4, 5
  )
)
column1 <- "tissue"
column2 <- "cluster"

head(df)
```

Implementing a greedy heuristic for weighted one-layer free problem
```{r}
clus_df_gather_sorted <- data_sort(df, column1="tissue", column2="cluster", sorting_algorithm="greedy_WOLF", fixed_column = "tissue")
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
alluvialmatch_objective_greedy_wolf <- determine_weighted_layer_free_objective(crossing_edges)
print(alluvialmatch_objective_greedy_wolf)
```

Plot the alluvial plot with sorting and color matching
```{r}
p1 <- plot_alluvial(clus_df_gather_sorted, column1="tissue", column2="cluster", column_weights = "value", sorting_algorithm="greedy_WOLF", fixed_column = "tissue", color_bands=TRUE)
p1
```

Repeat for all permutations of the right side, calculating objective for each one
```{r}
# Sample: assume clus_df_gather_sorted has factor 'cluster' with n levels
n <- nlevels(clus_df_gather_sorted$cluster)

# Get all n! permutations of labels 1 to n
perms <- gtools::permutations(n, n)

# Original cluster levels
orig_levels <- levels(clus_df_gather_sorted$cluster)

# Loop over permutations
objective_minimum = Inf
objectives <- numeric(nrow(perms))
for (i in 1:nrow(perms)) {
  new_labels <- perms[i, ]  # as.character(perms[i, ])
  
  # # Create a relabeled copy
  # clus_permuted <- clus_df_gather_sorted %>%
  #   mutate(cluster = factor(cluster, levels = orig_levels, labels = new_labels))

  map_dict <- setNames(new_labels, orig_levels)
  clus_permuted_objective <- determine_crossing_edges(clus_df_gather_sorted, column1 = "col1_int", column2 = "col2_int", fixed_column = 1, map_dict = map_dict, return_weighted_layer_free_objective = TRUE)
  
  objectives[i] <- clus_permuted_objective
  
  if (clus_permuted_objective < objective_minimum) {
      clus_permuted_best <- clus_df_gather_sorted %>%
        mutate(cluster = factor(cluster, levels = orig_levels, labels = new_labels))
  }
}
```

```{r}
clus_permuted_best
```


```{r}
p_best_wolf <- plot_alluvial(clus_permuted_best, column1="col1_int", column2="col2_int", column_weights = "value", sorting_algorithm="None", color_bands=TRUE)
p_best_wolf
```


```{r}
df_plot <- data.frame(
  perm_id = seq_along(objectives),
  objective = sort(objectives)
)

ggplot(df_plot, aes(x = perm_id, y = objective)) +
  geom_point() +
  geom_hline(yintercept = alluvialmatch_objective_greedy_wolf, linetype = "dashed", color = "blue") +
  annotate("text", x = Inf, y = alluvialmatch_objective_greedy_wolf, label = "alluvialmatch objective", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3) +
  labs(x = "Permutation index", y = "Objective value") +
  theme_minimal()

```

Now do the same for WBLF
```{r}
clus_df_gather_sorted <- data_sort(df, column1="tissue", column2="cluster", sorting_algorithm="greedy_WBLF", random_initializations = 10)
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
alluvialmatch_objective_greedy_wblf <- determine_weighted_layer_free_objective(crossing_edges)
print(alluvialmatch_objective_greedy_wblf)
```

```{r}
p2 <- plot_alluvial(clus_df_gather_sorted, column1="tissue", column2="cluster", column_weights = "value", sorting_algorithm="greedy_WBLF", color_bands=TRUE)
p2
```

Note that, above, we would reach the optimal if we moved up the cyan's 1 notch up

```{r}
# Get levels
left_levels <- levels(clus_df_gather_sorted$col1_int)
right_levels <- levels(clus_df_gather_sorted$col2_int)

n_left <- length(left_levels)
n_right <- length(right_levels)

# Get all permutations of each side
left_perms <- gtools::permutations(n_left, n_left)
right_perms <- gtools::permutations(n_right, n_right)

# Initialize tracking variables
objective_minimum <- Inf
objectives <- matrix(NA, nrow = nrow(left_perms), ncol = nrow(right_perms))
best_left_map <- NULL
best_right_map <- NULL

# Loop through all left and right permutations
for (i in 1:nrow(left_perms)) {
  left_new_labels <- left_perms[i, ]
  left_map <- setNames(left_new_labels, left_levels)
  
  for (j in 1:nrow(right_perms)) {
    right_new_labels <- right_perms[j, ]
    right_map <- setNames(right_new_labels, right_levels)

    # Apply mappings (e.g., to integer relabels)
    clus_permuted <- clus_df_gather_sorted %>%
      mutate(
        col1_int = left_map[as.character(col1_int)],
        col2_int = right_map[as.character(col2_int)]
      )

    # Compute objective
    obj <- determine_crossing_edges(
      clus_permuted,
      column1 = "col1_int",
      column2 = "col2_int",
      fixed_column = NULL,
      return_weighted_layer_free_objective = TRUE
    )

    objectives[i, j] <- obj

    # Update minimum if needed
    if (obj < objective_minimum) {
      objective_minimum <- obj
      clus_permuted_best <- clus_permuted
      best_left_map <- left_map
      best_right_map <- right_map
    }
  }
}

```

```{r}
p_best_wblf <- plot_alluvial(clus_permuted_best, column1="col1_int", column2="col2_int", column_weights = "value", sorting_algorithm="None", color_bands=TRUE)
p_best_wblf
```

Let's compare neighbornet too
```{r}
clus_df_gather_sorted_neighbornet <- data_sort(df, column1="tissue", column2="cluster", sorting_algorithm="neighbornet")
crossing_edges_output <- determine_crossing_edges(clus_df_gather_sorted_neighbornet, column1 = "col1_int", column2 = "col2_int")
crossing_edges <- crossing_edges_output$crossing_edges_df
neighbornet_objective_minimum <- determine_weighted_layer_free_objective(crossing_edges)
print(neighbornet_objective_minimum)

plot_alluvial(clus_df_gather_sorted_neighbornet, column1="tissue", column2="cluster", column_weights = "value", sorting_algorithm = "None", color_bands=TRUE)
```

```{r}
df_plot <- data.frame(
  perm_id = seq_along(objectives),
  objective = sort(objectives)
)

ggplot(df_plot, aes(x = perm_id, y = objective)) +
  geom_point() +
  geom_hline(yintercept = alluvialmatch_objective_greedy_wblf, linetype = "dashed", color = "blue") +
  annotate("text", x = Inf, y = alluvialmatch_objective_greedy_wblf, label = "alluvialmatch objective", 
           hjust = 1.1, vjust = -0.5, color = "blue", size = 3) +
  geom_hline(yintercept = neighbornet_objective_minimum, linetype = "dashed", color = "orange") +
  annotate("text", x = Inf, y = neighbornet_objective_minimum, label = "neighbornet objective", 
           hjust = 1.1, vjust = -0.5, color = "orange", size = 3) +
  labs(x = "Permutation index", y = "Objective value") +
  theme_minimal()
```

```{r}
sessioninfo::session_info()
```
